Exploring mathematical objects from custom-tailored mathematical universes

### 2020-01-20 paper-filmat.tex
Application of NSA in symplectic geometry

However, the realization of this fiction crucically rests on a non-principal
ultrafilter, whose existence requires principles which go beyond the means of
Zermelo--Fraenkel set theory~\textsc{zf}.\footnote{A hyperreal number is
represented by an infinite sequence~$(x_0,x_1,x_2,\ldots)$ of ordinary real
numbers. For instance, the sequence~$(1,1,1,\ldots)$ represents the hyperreal
version of the number~$1$, the sequence~$(1,\frac{1}{2},\frac{1}{3},\ldots)$
represents an infinitesimal number and its inverse~$(1,2,3,\ldots)$ represents
a transfinite number.
%
The sequence~$(1,1,1,\ldots)$ is deemed positive, and so is~$(-1,1,1,1,\ldots)$,
which differs from the former only in finitely many places. But
should~$(1,-1,1,-1,\ldots)$ be deemed positive or negative? Whatever the
answer, our decision has consequences for other sequences. For
instance~$(-1,1,-1,1,\ldots)$ should be assigned the opposite sign
and~$(\tan(1),\tan(-1),\tan(1),\tan(-1),\ldots)$ the same.
%
A non-principal ultrafilter is a set-theoretic gadget which fixes all such
decisions once and for all in a coherent manner. Having such an ultrafilter
available, a sequence~$(x_0,x_1,x_2,\ldots)$ is deemed positive if and only if
the set~$\{i \in \NN \,|\, x_i > 0\}$ is part of the ultrafilter.}
Non-principal ultrafilters cannot be described in explicit terms, and
they are also not at all canonical structures: \textsc{zfc} proves that there
are~$2^{2^{\aleph_0}}$ many~\cite{pospisil:ultrafilters}.

A practical consequence of this nonconstructivity is that it can be hard to
unwind proofs which employ hyperreal numbers to direct proofs, and even where
possible there is no general procedure for doing so.

@@

* XXX Syntactical vs. semantical interpretation
* XXX phone call analogy
* XXX quick overview of the several aspects of toposes (maybe at the end, as an outlook?)
* XXX allowing a switch of perspective
* XXX applications in mathematical practice
* XXX int. logic as common denominator
* XXX arbitrariness of the "standard axioms"
* XXX importance of having an adapted language
* XXX the internal language as a generalized modal operator
* XXX use the phrase "very rich and very deep" (or similarly) about set theory
* XXX Sven(?)
* XXX more pointers to topos literature
* XXX A topos is more than its semantics
* XXX Check references everywhere (give good pointers)
* XXX internal/external

>>>>

There is a growing body of research in mathematics which employs hyperreal
numbers in this sense. To exemplarily cite just one example, a recent
application of nonstandard analysis in symplectic geometry is due to Oliver
Fabert~\cite{fabert:non-squeezing,fabert:floer}, who verified an
infinite-dimensional analogue of the Arnold conjecture.

However, the realization of the fiction of infinitesimal numbers in nonstandard analysis crucically rests on a non-principal
ultrafilter, whose existence requires principles which go beyond the means of
Zermelo--Fraenkel set theory~\textsc{zf}.\footnote{A hyperreal number is
represented by an infinite sequence~$(x_0,x_1,x_2,\ldots)$ of ordinary real
numbers. For instance, the sequence~$(1,1,1,\ldots)$ represents the hyperreal
version of the number~$1$, the sequence~$(1,\frac{1}{2},\frac{1}{3},\ldots)$
represents an infinitesimal number and its inverse~$(1,2,3,\ldots)$ represents
a transfinite number.
%
The sequence~$(1,1,1,\ldots)$ is deemed positive, and so is~$(-1,1,1,1,\ldots)$,
which differs from the former only in finitely many places. But
should~$(1,-1,1,-1,\ldots)$ be deemed positive or negative? Whatever the
answer, our decision has consequences for other sequences. For
instance~$(-1,1,-1,1,\ldots)$ should be assigned the opposite sign
and~$(\tan(1),\tan(-1),\tan(1),\tan(-1),\ldots)$ the same.
%
A non-principal ultrafilter is a set-theoretic gadget which fixes all such
decisions once and for all in a coherent manner. Having such an ultrafilter
available, a sequence~$(x_0,x_1,x_2,\ldots)$ is deemed positive if and only if
the set~$\{i \in \NN \,|\, x_i > 0\}$ is part of the ultrafilter.}
Non-principal ultrafilters cannot be described in explicit terms, and
they are also not at all canonical structures: \textsc{zfc} proves that there
are~$2^{2^{\aleph_0}}$ many~\cite{pospisil:ultrafilters}.

A practical consequence of this nonconstructivity is that it can be hard to
unwind proofs which employ hyperreal numbers to direct proofs, and even where
possible there is no general procedure for doing so. For instance, Fabert has
not obtained a direct proof of his result, and not for the lack of trying
(personal communication).

@@

* XXX Syntactical vs. semantical interpretation
* XXX phone call analogy
* XXX quick overview of the several aspects of toposes (maybe at the end, as an outlook?)
* XXX allowing a switch of perspective
* XXX applications in mathematical practice
* XXX int. logic as common denominator
* XXX arbitrariness of the "standard axioms"
* XXX importance of having an adapted language
* XXX the internal language as a generalized modal operator
* XXX use the phrase "very rich and very deep" (or similarly) about set theory
* XXX more pointers to topos literature
* XXX A topos is more than its semantics
* XXX Check references everywhere (give good pointers)
* XXX internal/external
